{\bf Intrusion Detection using Parzen-Windows on Provenance Graph Statistics}

### 2013-12-10 writeup/paper-final.tex
Lots of copy editing

\begin{abstract}
Security rests on being able to identify potentially malicious behavior.
One way to identify malicious behavior is to compare system behavior to known normal behavior.
Intrusions on a system typically leave behind digital artifacts, and we hypothesize
that we can detect intrusions on a system by identifying anomalies in the artifacts' {\em provenance}, or the lineage of how
the artifact came to be in its present state.
We present an intrusion detection approach to find these anomalies by analyzing provenance graphs statistics. We use a Parzen-Window approach \cite{parzen} on various provenance graph statistics \cite{clustering} 
to determine probability density estimates of normal behavior, and we use these density estimates to 
determine if an intrusion occurred. We used this approach to analyze simple intrusions on small provenance graphs and achieved promising results, however applying this technique to {\em user-to-remote} (u2r) intrusions 
from the much larger 1998 DARPA Intrusion Detection data sets  \cite{darpa} resulted in normal and abnormal behavior being indistinguishable. We discuss why we received the results we obtained for each of these workloads, and we propose requirements on intrusion detection models using provenance graph models that are necessary for the model to scale to large provenance graphs.

@@

For as long as systems have existed, there have been malicious users that attempt to exploit vulnerabilities
in systems to gain unintended privileged access. As a result, system designers and 
administrators place a large effort in securing systems and preventing intrusions. However, intrusions
inevitably occur because of bugs or flaws within the system, and as a result, system administrators want to
have some automatic way of detecting these intrusions.

Intrusions often leave behind digital artifacts, e.g., unusual output or log files, a process executed
with an unusual arguments or environment, unusual system call activity by a process. In addition to
unusual digital artifacts, we hypothesize that
intrusions also leave behind abnormalities
in the {\em provenance} of these digital artifacts, the lineage of how these digital artifacts were created.

@@

Existing intrusion detection systems (IDS) use provenance data only to a limited extent.
Allen et al. analyze
basic statistics on provenance graphs (e.g., number of nodes, number of edges, max number of edges per node, etc.) in order to facilitate
easier manual intrusion detection by a human \cite{provstat}. Their work, however, does not present a way of using provenance
graphs to automatically determine intrusions. Other systems use provenance data to determine the scope of an
intrusion. For example, Backtracker \cite{backtracker} uses provenance graphs to build causality graphs of intrusions: once the
system has
determined an intrusion has occurred, the system follows the causality graph backwards determine the original source of the attack, and 
then it follows the causality graph forwards to determine all the objects tainted by the attack. However, the authors 
do not present a way of using provenance data to automatically detect intrusions.

\begin{figure}
  \label{graph-simple}
  \centering
    \includegraphics[width=0.5\textwidth]{img/graph-simple.png}
    \caption{Example of a simple provenance graph, showing the types of nodes and edges common in provenance graphs} 
\end{figure}

In this paper, we present an approach to use provenance data to automatically detect intrusions. Intrusion
detection can be framed as a {\em novelty detection} problem, in which one attempts to decide whether
an unknown test pattern is produced by an underlying distribution corresponding to a training set
of normal patterns \cite{parzen}. However, Yeung et al. note that in the case of intrusion detection, novel
or abnormal patterns are typically difficult to obtain, and as a result, they present a {\em Parzen-Window} approach that allows them to characterize abnormal behavior by only knowing normal behavior.
Using this technique, they obtain a high degree of success
in identifying various intrusion types \cite{parzen}. Because obtaining abnormal patterns is difficult, 
we borrow their Parzen-Window approach to build models of normal behavior using 
various provenance graph statistics and graph centrality metrics, and we evaluate the success of this technique
on {\em user-to-root} (u2r) intrusions, intrusions that provide unauthorized access to local superuser (root) privileges.

The main contributions of this paper are the following:
\begin{enumerate}
\item Discuss which provenance graph statistics we chose to use, how we chose them, and why we chose them.
\item Analyze the Parzen-Window technique with various provenance graph statistics on a small data set and a real intrusion detection data set and evaluate its accuracy on both.
\item Discuss the limitations of the approach and propose requirements for provenance graph intrusion detection models to scale to larger-sized graphs.
\end{enumerate}

\subsection{Outline}
In Section 2 we discuss work related to both provenance collection and intrusion detection. In Section 3 we detail our initial exploratory experiments on small data sets, determining which provenance graph features we wanted to explore further in designing our approach, as well as the implementation details of our initial experimental setup. In Section 4, we discuss the setup, results, and findings of our main experiment, in which we tested our approach on existing large intrusion detection data sets. In Section 5 we discuss future work and directions that this project has motivated, and in Section 6 we conclude.

@@

Many existing intrusion detection systems use provenance in some limited capacity; however, none of them
have used provenance for a fully automated detection system. 
Somayaji and Forrest's work in analyzing sequences of system calls for intrusion detection \cite{somayaji, somayaji-recent} can
be seen as one of the earliest works in building an intrusion detection system using provenance: sequences of
system calls can be seen as a limited form of provenance, as many systems build provenance graphs from
system call traces \cite{spade}. By analyzing the sequences
of system calls a process makes over time, their system builds a profile of ``normal" process behavior. When
the process in the future makes enough patterns of unrecognized sequences of system calls, the system flags
this process as behaving abnormally and attempts to stop the process, either through exponentially slowing down
system calls or aborting system calls entirely. Notably, their system is trained only by analyzing ``normal" data
without knowing explicitly what is considered ``abnormal." Because their system achieves high
accuracy in determining intrusions with only sequences of system calls, we believe that having real provenance
data in a graph structure--which in theory would be a superset of system call data--would allow us
to achieve comparable, if not better intrusion detection accuracy.

@@

Allen et al. make use of simple provenance graph statistics to categorize provenance graphs and to assist a human
in manual intrusion-detection \cite{provstat}. Given a data set of many provenance graphs, they calculated basic statistics
on the graphs (e.g., total nodes, total edges, max incoming edges on a node, max outgoing edges on a node,
node/edge ratio, average total edges per node, etc.). Their approach allows a human to more easily
notice anomalies in provenance graphs, but they do not provide a way to automate this detection.
We borrow their ideas of using simple provenance graph statistics as features to detect intrusions.

@@

To address the difficulty of obtaining abnormal patterns for novelty detection on intrusion detection, we borrow
the Parzen-Window model approach proposed by Yeung et al.
They attempt to solve the lack of abnormal patterns problem by using Parzen-Windows to build {\em nonparametric}
density estimations of normal behavior \cite{parzen}. A density estimation is considered {\em nonparametric} if the estimation
makes no assumptions about the forms of the PDFs, except that PDFs are smooth. By using a nonparametric
density estimation, one can build a probability density estimate of normal behavior by only having examples of normal
behavior and no examples of abnormal behavior.

@@

In this paper, we use this same model to build profiles of ``normal" behavior for each process. We use
one-dimensional feature vectors $\m{x}$, and we use various provenance graph statistics and graph
centrality metrics as our features. We describe the various statistics and graph centrality metrics we explored
in the following section. For the common variance of our Gaussian kernels, we chose
the variance to be the variance of the data set $D_P$ for each unique process name $P$. 
We vary the features we used and values of $\psi$ and evaluate its performance on an intrusion detection data set.

@@

Furthermore, we hypothesize that intrusions manifest as anomalies in the lineage of digital artifacts, and
one way of observing this is through {\em graph centrality metrics}. A {\em graph centrality metric} is a function
of a node and its containing graph, and intuitively represents how {\em central} or {\em important} the node
is in the graph. In terms of provenance graphs, processes of the same program typically behave similarly to one another
and this should result in similar number of inputs, outputs, number of ancestors, and number of descendants for normal
process nodes of the same program.
Intrusions typically change the number of expected inputs and outputs (e.g., a process forking a shell process
when it normally does not), and we hypothesize that this can manifest as unusual changes in centrality
for the offending node in the provenance graph. Macko et al. note the following 3 observations unique to provenance graphs \cite{clustering}:
\begin{enumerate}
\item The ubiquity (importance) of a node is a function of a node's descendants, not its ancestors. (Note: Because edges in provenance graphs are in the direction of data dependence instead of data flow, ``descendants" in this case means ancestors in the provenance graph).
\item There is no agreed-upon granularity in which provenance should be captured.
\item Provenance has a temporal component.
\end{enumerate}

Using these observations, they propose several centrality metrics on provenance graphs:
\begin{itemize}
\item \textbf{In-degree Centrality}. Because a node's importance is based on its output descendants, then the number of times a node was used as input to another node indicates a simple measure of importance.
\item \textbf{Age}. They note that large "jumps" in timestamps or age might indicate breaks between tasks, but this interpretation is complicated when scripts execute tasks in short succession.
\item \textbf{Ancestor Centrality}. This measures the total number of descendants of a node, normalized by the total number of nodes in a graph.
\item \textbf{Opsahl's Closeness Centrality}. This accounts for the number of edges between a node and its descendants. To compute this, we calculate
$$CC'(v) = \sum_{x \in V -  \{v\}} (d'_{xv})^{-1}$$
where $V$ is the set of nodes in the graph, $d'_{xv}$ is the distance from node $x$ to $v$ by only following outgoing edges from $x$. Unreachable nodes from $x$ will have $d'_{xv} = \infty$, and so $(d'_{xv})^{-1} = 0$.
\item \textbf{Provenance Eigenvector Centrality}. The centrality of the nodes is given by the left dominant eigenvector of the matrix:
$$M_{ij} =
\begin{cases}
1 & \text{if there is an edge } i \to j \\
1/|V| & \text{if there is no out edge from } i \\
0 & \text{otherwise}
\end{cases}
$$
\end{itemize}
We implemented some of these graph centrality metrics and used Parzen-Windows on centrality measurements to compute density estimates for each process.

@@

We envision a provenance aware system which constantly builds normal profiles of processes. Simultaneously,
the system looks for instances of deviation from normal behavior and handles them appropriately
(e.g., terminating the process, slowing down the process). In our work, we derive normal profiles of process behavior
from provenance graphs, and we use a Parzen-Window model on various provenance graph statistics
to determine if anomalous nodes appear in the graph.

@@

To determine which provenance statistics seem useful, we first created simple simulated intrusions
and ran them in a Provenance Aware Storage System (PASS) \cite{pass}, a file system with
a mounted volume that automatically collects provenance of all files within that volume.
To make data collection easier, we used PASSv2 \cite{passv2}, which provides an Ubuntu virtual machine with the PASS volume already installed.

For each of our intrusions, we ran it normally multiple times and collected the provenance. We then ran the
intrusion with the exploit multiple times and collected the new provenance.

We used the following procedure to implement our Parzen-Window density estimates and to decide whether a node is behaving normally or abnormally.
\begin{enumerate}
\item For each statistic that we implemented, we computed the counts for that statistic for all the
nodes in the provenance graph containing the normally-behaving workload.
\item We aggregated the counts by the name of the process (the path to the executable). For a given process name $P$, denote the
set of counts for that process name as $D_P$.
\item For each $y \in D_P$, calculate the log-likelihood $L_P(y)$.
\item For a test node with process name $P$ in the exploited-workload provenance graph, we calculate the value $x$ of the statistic in its graph, and then compute $L_P(x)$. 
If the ratio of the number of nodes $y$ with $L_P(y) \leq L_P(x)$,
to the total number of nodes in $D_P$ is greater than the false detection rate $\psi$, we label the node as normal. Otherwise we label the node as
an intrusion.
\end{enumerate}

@@

We ran \texttt{hello} normally 40 times on various inputs and collected the provenance times. We then exploited the buffer overflow vulnerability in \texttt{hello} to spawn an \texttt{ls} process, and saved the standard out of that process to a file. See Figure 2 for examples of provenance graphs for the normally behaving and exploited versions of the program.

@@

We ran our program on over 200 good inputs, and stored the provenance data. We then ran our program on two bad inputs, first to call \texttt{ls} as a Proof of Concept, and again to call \texttt{/bin/sh} to demonstrate that more severe attacks can be performed. We again exported the provenance generated from this exploited workload. See Figure 3 for examples of provenance graphs for the normally behaving and exploited versions of the program.

@@

For our system administration attack, we compiled the PASS toolset using \texttt{gcc}, and exported the provenance from these runs. We then replaced \texttt{gcc} with our exploited version, compiled the toolset several times with different configurations, and exported the provenance again.

@@

In Table 1, we show a summary of the statistics that had any statistical difference between normal and abnormally behaving nodes, and Figure 4 highlights the most salient statistics for the \texttt{gcc} exploit. The buffer overflows showed the most abnormal behavior in the number of outgoing processes and the centrality measures. This makes sense, as arbitrary code execution will generate additional processes. Of the centrality metrics we implemented, the Opsahl Centrality metric did the best in distinguishing an exploited node from normally behaving nodes. Because the Opsahl Centrality metric accounts for the number of edges between a node and all of its descendants, this centrality metric is very sensitive to the addition of nodes. Exploits that have additional outputs (e.g. new processes and output files through a remote shell), thus generate a higher Opsahl centrality metric for the offending node. The more active the exploit is (say, if it forks a user shell), the more the centrality measure will be impacted, and the more likely it is that our model will catch the exploit.

@@

For our experiments, we limited the scope of our intrusion detection system to only u2r attacks, as
the other kinds of attacks were network-based and would not appear in the provenance data. 
We identify each workload with a tuple:
$$(\psi, W)$$
where $\psi$ is our false detection rate for our Parzen-Window
estimates, and $W$ is the time window that we use to form time intervals to look for anomalous nodes. We used
the following procedure to evaluate the accuracy of our system: 
\begin{enumerate}
\item Use BSM traces to build provenance graphs, in a similar fashion to SPADE \cite{spade}. 
\item Compute the Opsahl centrality of the nodes in the graph and create a mapping from each process
name $P$ to a list of log-likelihood values $L(y)$, where $y$ is the computer centrality measurement for each node in the graph having process name $P$.
\item For a specific moment in time $T$, find all edges in the provenance graph we built with timestamps in the interval $[T - W, T + W]$. 
Use the nodes attached to these edges as our {\em candidate anomaly set}. We used the specific times listed in data sets for our values of $T$.
\item For all nodes in the candidate anomaly set, find the process name $P$ of the node, and use our Parzen-Window approach on process
name $P$ to determine if the node is anomalous. We evaluate whether a node is normal or anomalous using the Parzen-Windows built from the graph
containing the node.
\item For times $T$ in the data set listed as having an intrusion, our intrusion detection system considers that time $T$ as having an intrusion
if a {\em majority} of the nodes in the candidate anomaly set are considered anomalous.
\end{enumerate}

@@

We evaluated the accuracy of our system on various values of $\psi$ and $W = 10s, 100s, 1000s$.
\subsection{BSM to Provenance Conversion}
For our experiment, we needed to convert the BSM traces into a provenance graph. The designers of SPADE \cite{spade} used this same approach, so we ported their conversion code to our system. The BSM data only contains PIDs, however, not process names. While SPADE used \texttt{ps} to determine the process names for a particular PID, we needed to find another way, because we did not have access to the \texttt{ps} output for all timestamps. We initially tried to use minutely \texttt{ps} output provided by DARPA, but too many processes started and completed within this window, leaving them unnamed in our graph. We then parsed the BSM traces directly, looking for \texttt{execve} calls, which we could use to name our process nodes.

The SPADE code we used as a model parses the BSM file, adding nodes when processes and files are introduced (e.g. via \texttt{fork} or \texttt{creat}) and adding edges when processes and files interact (e.g. via \texttt{write} and \texttt{fork}). The types of edges were very similar to those found in the PASS data - namely ``triggered by'' edges between process nodes and ``used'' edges between process nodes and file nodes.
\subsection{Results}

\noindent
\begin{figure}
  \label{tpr}
  \centering
    \includegraphics[width=0.5\textwidth]{img/tpr.png}
    \caption{True Positive Rate. The window size 10s, 100s, 1000s did not impact the true positive rate at all for various values of $\psi$.} 
\end{figure}

\noindent
\begin{figure}
  \label{fpr}
  \centering
    \includegraphics[width=0.5\textwidth]{img/fpr.png}
    \caption{False Positive Rate. The window size 10s, 100s, 1000s did not impact the false positive rate very
    much for various values of $\psi$.} 
\end{figure}

We plot the true positive rates for various values of $\psi$ and window sizes $W$ in Figure 5, and similarly for the false positive rates in Figure 6.

@@

We believe that the reason we had more promising results on our simulated intrusion data sets from PASS was because the provenance graphs from these workloads were much smaller (on the order of 5000 nodes). As a result of the much smaller size, small changes in provenance graph structure from an intrusion has a much profounder impact on the smaller provenance graphs. Thus, intrusions on our PASS workload had a much higher impact on the centrality calculations and resulted in more anomalous behavior in our Parzen-Window model. In the DARPA data set, we faced much larger provenance graphs (on the order of 20,000 nodes), and because of the much larger size, the effect of small changes in provenance graph structure from an intrusion had a much smaller impact on the overall structure of the graph. Intrusions did not change the centrality metrics very much in large graphs; therefore anomalous behavior went undetected, regardless of the window size we chose. Since our system could not distinguish between anomalous and normal behavior, then the true positive rate and false positive rate are essentially identical. In addition, we found that many more nodes were being flagged as intrusions (indeed, a positive result meant that a majority of nodes in a given time range were considered anomalous). In Figure 7, we see that the percentage of nodes during exploit windows were identified similarly to those during windows containing no exploit. As expected, as PSI increases, so too does the number of events marked as exploits. This result differed greatly from our preliminary exploration, where normal nodes were consistently identified as normal. We believe this occurred because the DARPA dataset came from a real system running for days, so the variety of nodes increased, and the distribution of these metrics became more full, so complete outliers were less present. 
\begin{figure}
  \label{intrusion-v-psi}
  \centering
    \includegraphics[width=0.5\textwidth]{img/intrusion_v_psi.png}
    \caption{The percentage of nodes flagged as an intrusion for various values of PSI. This figure was generated from the Week 5 Monday data, using a window size of 10 seconds. Other window sizes showed similar results. Recall that our system flags events with a percentage of 0.5 or greater as ``anomalous''.} 
\end{figure}
From these results, we propose the following requirements for intrusion detection models that use provenance graph statistics in order to be useful for large provenance graphs:

>>>>

\begin{abstract}
Security depends on being able to identify potentially malicious behavior.
One way to identify malicious behavior is to compare system behavior to known normal behavior.
Intrusions on a system typically leave behind digital artifacts, and we hypothesize
that we can detect intrusions on a system by identifying anomalies in the artifacts' {\em provenance}, or the lineage of how
the artifact came to be in its present state.
We present an intrusion detection approach to find these anomalies by analyzing provenance graph statistics. We use a Parzen-Window approach \cite{parzen} on various provenance graph statistics \cite{clustering} 
to determine probability density estimates of normal behavior, and we use these density estimates to 
determine if an intrusion occurred. We used this approach to analyze simple intrusions on small provenance graphs and achieved promising results. However, applying this technique to {\em user-to-remote} (u2r) intrusions 
from the much larger 1998 DARPA Intrusion Detection data sets  \cite{darpa} resulted in normal and abnormal behavior being indistinguishable. We discuss why we received the results we obtained for each of these workloads, and we propose requirements on intrusion detection models using provenance graph models that are necessary for the model to scale to large provenance graphs.

@@

For as long as systems have existed, there have been malicious users that attempt to exploit vulnerabilities
in systems to gain unintended privileged access. As a result, system designers and 
administrators make a large effort to secure systems and prevent intrusions. However, intrusions
inevitably occur because of bugs or flaws within the system, and as a result, system administrators want to
have some automatic way of detecting these intrusions.

Intrusions often leave behind digital artifacts, e.g., unusual output or log files, a process executed
with unusual arguments or environment, or unusual system call activity by a process. In addition to
unusual digital artifacts, we hypothesize that
intrusions also leave behind abnormalities
in the {\em provenance} of these digital artifacts, the lineage of how these digital artifacts were created.

@@

Existing intrusion detection systems (IDS) use provenance data only to a limited extent.
Allen et al. analyze
basic statistics on provenance graphs (e.g., number of nodes, number of edges, max number of edges per node, etc.) in order to facilitate
easier manual intrusion detection by a human \cite{provstat}. Their work, however, does not present a way of using provenance
graphs to automatically determine whether intrusions have occurred. Other systems use provenance data to determine the scope of an
intrusion. For example, Backtracker \cite{backtracker} uses provenance graphs to build causality graphs of intrusions: once the
system has
determined an intrusion has occurred, the system follows the causality graph backwards to determine the original source of the attack, and 
then it follows the causality graph forward to determine all the objects tainted by the attack. However, the authors 
do not present a way of using provenance data to automatically detect intrusions.

\begin{figure}
  \label{graph-simple}
  \centering
    \includegraphics[width=0.5\textwidth]{img/graph-simple.png}
    \caption{Example of a simple provenance graph, showing the types of nodes and edges common in provenance graphs.} 
\end{figure}

In this paper, we present an approach to use provenance data to automatically detect intrusions. Intrusion
detection can be framed as a {\em novelty detection} problem, in which one attempts to decide whether
an unknown test pattern is produced by an underlying distribution corresponding to a training set
of normal patterns \cite{parzen}. However, Yeung et al. note that in the case of intrusion detection, novel
or abnormal patterns are typically difficult to obtain, and as a result, they present a {\em Parzen-Window} approach that allows them to characterize abnormal behavior using only normal behavior.
With this technique, they obtain a high degree of success
in identifying various intrusion types \cite{parzen}. Because obtaining abnormal patterns is difficult, 
we borrow their Parzen-Window approach to build models of normal behavior using 
various provenance graph statistics and graph centrality metrics, and we evaluate the success of this technique
on {\em user-to-root} (u2r) intrusions, intrusions that provide unauthorized access to local superuser (root) privileges.

The main contributions of this paper are the following:
\begin{enumerate}
\item Discuss which provenance graph statistics we chose to use, how we chose them, and why we chose them.
\item Analyze the Parzen-Window technique with various provenance graph statistics on a small data set and a real intrusion detection data set and evaluate the accuracy we achieve for both.
\item Discuss the limitations of the approach and propose requirements for provenance graph intrusion detection models to scale to larger-sized graphs.
\end{enumerate}

\subsection{Outline}
In Section 2, we discuss work related to both provenance collection and intrusion detection. In Section 3, we detail our initial exploratory experiments on small data sets, determining which provenance graph features we wanted to explore further in designing our approach, as well as the implementation details of our initial experimental setup. In Section 4, we discuss the setup, results, and findings of our main experiment, in which we tested our approach on existing large intrusion detection data sets. In Section 5, we discuss future work and directions that this project has motivated, and in Section 6 we conclude.

@@

Many existing intrusion detection systems use provenance in some limited capacity; however, none of them
have used provenance for a fully automated detection system. 
Somayaji and Forrest's work in analyzing sequences of system calls for intrusion detection \cite{somayaji, somayaji-recent} can
be seen as one of the earliest works in building an intrusion detection system using provenance: sequences of
system calls can be seen as a limited form of provenance, as many systems build provenance graphs from
system call traces (as done in SPADE \cite{spade}). By analyzing the sequences
of system calls a process makes over time, their system builds a profile of ``normal" process behavior. When
the process in the future makes enough patterns of unrecognized sequences of system calls, the system flags
this process as behaving abnormally and attempts to stop the process, either through exponentially slowing down
system calls or aborting system calls entirely. Notably, their system is trained only by analyzing ``normal" data
without knowing explicitly what is considered ``abnormal." Because their system achieves high
accuracy in determining intrusions with only sequences of system calls, we believe that having real provenance
data in a graph structure--which in theory would be a superset of system call data--would allow us
to achieve comparable, if not better intrusion detection accuracy.

@@

Allen et al. make use of simple provenance graph statistics to categorize provenance graphs and to assist a human
in manual intrusion detection \cite{provstat}. Given a data set of many provenance graphs, they calculated basic statistics
of the graphs (e.g., total nodes, total edges, max incoming edges on a node, max outgoing edges on a node,
node/edge ratio, average total edges per node, etc.). Their approach allows a human to more easily
notice anomalies in provenance graphs, but they do not provide a way to automate this detection.
We borrow their ideas of using simple provenance graph statistics as features to detect intrusions.

@@

To address the difficulty of obtaining abnormal patterns for intrusion detection, we borrow
the Parzen-Window model approach proposed by Yeung et al.
They attempt to solve the lack of abnormal patterns problem by using Parzen-Windows to build {\em nonparametric}
density estimations of normal behavior \cite{parzen}. A density estimation is considered {\em nonparametric} if the estimation
makes no assumptions about the forms of the PDFs, except that PDFs are smooth. By using a nonparametric
density estimation, one can build a probability density estimate of normal behavior by only having examples of normal
behavior and no examples of abnormal behavior.

@@

In this paper, we use this same model to build profiles of ``normal" behavior for each process. We use
one-dimensional feature vectors $\m{x}$, and we use various provenance graph statistics and graph
centrality metrics as our features. In the following section, we describe the various statistics and graph centrality metrics we explored. For the common variance of our Gaussian kernels, we chose
the variance to be the variance of the data set $D_P$ for each unique process name $P$. 
We vary the features we used and values of $\psi$ and evaluate its performance on an intrusion detection data set.

@@

Furthermore, we hypothesize that intrusions manifest as anomalies in the lineage of digital artifacts, and
one way of observing this is through {\em graph centrality metrics}. A {\em graph centrality metric} is a function
of a node and the graph that contains it, and the value of the function intuitively represents how {\em central} or {\em important} the node
is in the graph. In terms of provenance graphs, processes of the same program typically behave similarly to one another
and this should result in similar number of inputs, outputs, number of ancestors, and number of descendants for normal
process nodes of the same program.
Intrusions typically change the number of expected inputs and outputs (e.g., a process forking a shell process
when it normally does not), and we hypothesize that this can manifest as unusual changes in centrality
for the offending node in the provenance graph. Macko et al. note the following 3 observations unique to provenance graphs \cite{clustering}:
\begin{enumerate}
\item The ubiquity (importance) of a node is a function of a node's descendants, not its ancestors. (Note: Because edges in provenance graphs are in the direction of data dependence instead of data flow, ``descendants" in this case means ancestors in the provenance graph).
\item There is no agreed-upon granularity in which provenance should be captured.
\item Provenance has a temporal component.
\end{enumerate}

Using these observations, they propose several centrality metrics on provenance graphs:
\begin{itemize}
\item \textbf{In-degree Centrality}. Because a node's importance is based on its output descendants, then the number of times a node was used as input to another node indicates a simple measure of importance.
\item \textbf{Age}. They note that large ``jumps'' in timestamps or age might indicate breaks between tasks, but this interpretation is complicated when scripts execute tasks in short succession.
\item \textbf{Ancestor Centrality}. This metric measures the total number of descendants of a node, normalized by the total number of nodes in a graph.
\item \textbf{Opsahl's Closeness Centrality}. This metric accounts for the number of edges between a node and its descendants. To compute this, we calculate
$$CC'(v) = \sum_{x \in V -  \{v\}} (d'_{xv})^{-1}$$
where $V$ is the set of nodes in the graph, $d'_{xv}$ is the distance from node $x$ to $v$ by only following outgoing edges from $x$. Unreachable nodes from $x$ will have $d'_{xv} = \infty$, and so $(d'_{xv})^{-1} = 0$.
\item \textbf{Provenance Eigenvector Centrality}. The centrality of the nodes is given by the left dominant eigenvector of the matrix:
$$M_{ij} =
\begin{cases}
1 & \text{if there is an edge } i \to j \\
1/|V| & \text{if there is no out edge from } i \\
0 & \text{otherwise}
\end{cases}
$$
\end{itemize}
We implemented some of these graph centrality metrics and used Parzen-Windows on centrality measurements to compute density estimates for each process.

@@

We envision a provenance aware system which constantly builds normal profiles of processes. Simultaneously,
the system looks for instances of deviation from normal behavior and handles them appropriately
(e.g., terminating the process or slowing down the process). In our work, we derive normal profiles of process behavior
from provenance graphs, and we use a Parzen-Window model on various provenance graph statistics
to determine if anomalous nodes appear in the graph.

@@

To determine which provenance statistics seem useful, we first created simple simulated intrusions
and ran them in a Provenance Aware Storage System (PASS) \cite{pass}, a file system with
a mounted volume that automatically collects provenance of all files within that volume.
To make data collection easier, we used PASSv2 \cite{passv2}, which provides an Ubuntu virtual machine with a PASS volume already installed.

We ran each of our exploited programs normally (i.e. without taking advantage of the exploit) multiple times and collected the provenance. We then ran the
intrusion with the exploit multiple times and collected the new provenance.

We used the following procedure to implement our Parzen-Window density estimates and to decide whether a node is behaving normally or abnormally.
\begin{enumerate}
\item For each statistic that we implemented, compute the counts for that statistic for all the
nodes in the provenance graph containing the normally-behaving workload.
\item Aggregated the counts by the name of the process (the path to the executable). For a given process name $P$, denote the
set of counts for that process name as $D_P$.
\item For each $y \in D_P$, calculate the log-likelihood $L_P(y)$.
\item For a test node with process name $P$ in the exploited-workload provenance graph, calculate the value $x$ of the statistic in its graph, and then compute $L_P(x)$. 
If the ratio of the number of nodes $y$ with $L_P(y) \leq L_P(x)$,
to the total number of nodes in $D_P$ is greater than the false detection rate $\psi$, label the node as normal. Otherwise label the node as
an intrusion.
\end{enumerate}

@@

We ran \texttt{hello} normally 40 times on various inputs and collected the provenance. We then exploited the buffer overflow vulnerability in \texttt{hello} to spawn an \texttt{ls} process and saved the standard out of that process to a file. See Figure 2 for examples of provenance graphs for the normally behaving and exploited versions of the program.

@@

We ran our program on over 200 good inputs and stored the provenance data. We then ran our program on two bad inputs, first to call \texttt{ls} as a Proof of Concept, and again to call \texttt{/bin/sh} to demonstrate that more severe attacks can be performed. We again exported the provenance generated from this exploited workload. See Figure 3 for examples of provenance graphs for the normally behaving and exploited versions of the program.

@@

For our system administration attack, we compiled the PASS toolset using \texttt{gcc} and exported the provenance from these runs. We then replaced \texttt{gcc} with our exploited version, compiled the toolset several times with different configurations, and exported the provenance again.

@@

In Table 1, we show a summary of the statistics that had any statistical difference between normally and abnormally behaving nodes. Figure 4 highlights the most salient statistics for the \texttt{gcc} exploit. The buffer overflows showed the most abnormal behavior in the number of outgoing processes and the centrality measures. This makes sense, as arbitrary code execution will generate additional processes. Of the centrality metrics we implemented, the Opsahl Centrality metric performed best in distinguishing an exploited node from normally behaving nodes. Because the Opsahl Centrality metric accounts for the number of edges between a node and all of its descendants, this centrality metric is very sensitive to the addition of nodes. Exploits that have additional outputs (e.g. new processes and output files through a remote shell), thus generate a higher Opsahl centrality metric for the offending node. The more active the exploit is (say, if it forks a user shell), the more the centrality measure will be impacted, and the more likely it is that our model will catch the exploit.

@@

For our experiments, we limited the scope of our intrusion detection system to only u2r attacks, as
the other kinds of attacks were network-based and would not appear in the provenance data. 
We identify each workload with a tuple:
$$(\psi, W)$$
where $\psi$ is our false detection rate for our Parzen-Window
estimates, and $W$ is the time window that we use to form time intervals to look for anomalous nodes. We used
the following procedure to evaluate the accuracy of our system: 
\begin{enumerate}
\item Use BSM traces to build provenance graphs, in a similar fashion to SPADE \cite{spade}. 
\item Compute the Opsahl centrality of the nodes in the graph and create a mapping from each process
name $P$ to a list of log-likelihood values $L(y)$, where $y$ is the computer centrality measurement for each node in the graph having process name $P$.
\item For a specific moment in time $T$, find all edges in the provenance graph we built with timestamps in the interval $[T - W, T + W]$. 
Use the nodes attached to these edges as our {\em candidate anomaly set}. We used the specific times listed in data sets for our values of $T$.
\item For all nodes in the candidate anomaly set, find the process name $P$ of the node, and use our Parzen-Window approach on process
name $P$ to determine if the node is anomalous. Evaluate whether a node is normal or anomalous using the Parzen-Windows built from the graph
containing the node.
\item For times $T$ in the data set listed as having an intrusion, our intrusion detection system considers that time $T$ as having an intrusion
if a {\em majority} of the nodes in the candidate anomaly set are considered anomalous.
\end{enumerate}

@@

We evaluated the accuracy of our system on various values of $\psi$ and $W = 10s, 100s, 1000s$.

\subsection{BSM to Provenance Conversion}
For our experiment, we needed to convert the BSM traces into a provenance graph. The designers of SPADE \cite{spade} used this same approach, so we ported their conversion code to our system. The BSM data only contains PIDs, however, not process names. While SPADE used \texttt{ps} to determine the process names for a particular PID, we needed to find another way, because we did not have access to the \texttt{ps} output for all timestamps. We initially tried to use \texttt{ps} output provided by DARPA, but they only provided data at 60 second intervals, and too many processes started and completed within this window, leaving them unnamed in our graph. We then parsed the BSM traces directly, looking for \texttt{execve} calls, which we could use to name our process nodes.

The SPADE code we used as a model parses the BSM file, adding nodes when processes and files are introduced (e.g. via \texttt{fork} or \texttt{creat}) and adding edges when processes and files interact (e.g. via \texttt{write} and \texttt{fork}). The types of edges were very similar to those found in the PASS data - namely, ``triggered by'' edges between process nodes and ``used'' edges between process nodes and file nodes.
\subsection{Results}

\noindent
\begin{figure}
  \label{tpr}
  \centering
    \includegraphics[width=0.5\textwidth]{img/tpr.png}
    \caption{True Positive Rate. The window size 10s, 100s, 1000s did not impact the true positive rate in any way for various values of $\psi$.} 
\end{figure}

\noindent
\begin{figure}
  \label{fpr}
  \centering
    \includegraphics[width=0.5\textwidth]{img/fpr.png}
    \caption{False Positive Rate. The window size 10s, 100s, 1000s did not impact the false positive rate significantly for various values of $\psi$.} 
\end{figure}

We plot the true positive rates for various values of $\psi$ and window sizes $W$ in Figure 5 and similarly for the false positive rates in Figure 6.

@@

We believe that the reason we had more promising results on our simulated intrusion data sets from PASS was because the provenance graphs from these workloads were much smaller (on the order of 5,000 nodes). As a result of the much smaller size, small changes in provenance graph structure from an intrusion has a much profounder impact on the smaller provenance graphs. Thus, intrusions on our PASS workload had a much higher impact on the centrality calculations and resulted in more anomalous behavior in our Parzen-Window model. In the DARPA data set, we faced much larger provenance graphs (on the order of 20,000 nodes), and because of the much larger size, the effect of small changes in provenance graph structure from an intrusion had a much smaller impact on the overall structure of the graph. Intrusions did not change the centrality metrics very much in large graphs; therefore anomalous behavior went undetected, regardless of the window size we chose. Since our system could not distinguish between anomalous and normal behavior, the true positive rates and false positive rates are essentially identical. In addition, we found that many more nodes were being flagged as intrusions (indeed, a positive result meant that a majority of nodes in a given time range were considered anomalous). In Figure 7, we see that the percentage of nodes during exploit windows that were identified as intrusions is very similar to the percentage identified during windows containing no exploit. As expected, as $\psi$ increases, so too does the number of events marked as exploits. This result differed greatly from our preliminary exploration, where normal nodes were consistently identified as normal. We believe this occurred because the DARPA dataset came from a real system running for days, so the variety of nodes increased, and the distribution of these metrics became more full, so the statistics contained fewer complete outliers. 
\begin{figure}
  \label{intrusion-v-psi}
  \centering
    \includegraphics[width=0.5\textwidth]{img/intrusion_v_psi.png}
    \caption{The percentage of nodes flagged as intrusions for various values of $\psi$. This figure was generated from the Week 5 Monday DARPA data, using a window size of 10 seconds. Other window sizes showed similar results. Recall that our system flags events with a percentage of 0.5 or greater as ``anomalous''.} 
\end{figure}
From these results, we propose the following requirements for intrusion detection models that use provenance graph statistics in order to be useful for large provenance graphs:
